---
title: "2D Quantitative Data: Scatterplots and Linear Regression"
author: "Prof Ron Yurko"
footer:  "[statds-36315-spring25](https://ryurko.github.io/statds-36315-spring25/)"
date: 2025-02-17
engine: knitr
format:
  revealjs:
    theme: theme.scss
    chalkboard: true
    pdf-separate-fragments: true
    slide-number: c/t
    smaller: true
    code-line-numbers: true
    linestretch: 1.25
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(tidyverse)
library(palmerpenguins)
```

## Announcements, previously, and today...

::: {style="font-size: 60%;"}

**HW4 is due Wednesday by 11:59 PM and you have Lab 6 again on Friday!**

**Take-home exam is next week Wednesday Feb 26th**

Here's how the exam will work:

+ I'll post the exam Monday evening, and it's due Wednesday by 11:59 PM EDT (Feb 26th)
  
+ Exam will cover material from HW 1-4 and Labs 1-6
  
+ Questions will be similar to homeworks but more open-ended, e.g, instead of "make a side-by-side violin plot..." I'll ask "Make a plot that compares the conditional distributions..."
  
+ __There will NOT be class on Wednesday Feb 26th__ 
  
+ Conflict Feb 26th? __Let me know ASAP!__ Day-of accommodations will NOT be made, late submissions will NOT be accepted

:::

. . .

::: {style="font-size: 60%;"}


Discussed power in the context of visualizations and statistical tests

Even if there is a true effect, you may have limited power to detect it

Several ways to formally compare distributions:

* $t$-test: Compare means

* Bartlett's test: Compare variances

* KS test: Compare distributions


:::

. . .

::: {style="font-size: 80%;"}

**TODAY:** 2D quantitative data, scatterplots, and linear regression

:::

---

## 2D quantitative data

- We're working with two variables: $(X, Y) \in \mathbb{R}^2$, i.e., dataset with $n$ rows and 2 columns

. . .

- Goals:

  - describing the relationships between two variables
  
  - describing the conditional distribution $Y | X$ via regression analysis
  
  - describing the joint distribution $X,Y$ via contours, heatmaps, etc.
  
. . .

- Few big picture ideas to keep in mind:

  - scatterplots are by far the most common visual
  
  - regression analysis is by far the most popular analysis (you have a whole class on this...)
  
  - relationships may vary across other variables, e.g., categorical variables
  
---

## Making scatterplots with `geom_point()`

```{r}
#| code-line-numbers: "2-3"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```


---

## ALWAYS adjust the `alpha`

```{r}
#| code-line-numbers: "3"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.5)
```

---

## Map categorical variable to `color`

```{r}
#| code-line-numbers: "3"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = species)) +
  geom_point(alpha = 0.5)
```

---

## Map continuous variable to `color`

```{r}
#| code-line-numbers: "3"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = body_mass_g)) +
  geom_point(alpha = 0.5)
```

---

## Map continuous variable to `color`

```{r}
#| code-line-numbers: "5"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = body_mass_g)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "darkblue", high = "darkorange")
```

---

## Map continuous variable to `size`

```{r}
#| code-line-numbers: "3"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             size = body_mass_g)) +
  geom_point(alpha = 0.5)
```

---

## Map categorical variable to `shape`


```{r}
#| code-line-numbers: "3"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             shape = species)) +
  geom_point(alpha = 0.5)
```

---

## ALL AT ONCE!

```{r}
penguins |>
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm,
             color = species, shape = island, size = body_mass_g)) +
  geom_point(alpha = 0.5) 
```


---

![](https://images.squarespace-cdn.com/content/v1/5ce45e9df03bfe000172d138/1597870238202-3OWP5306LOI8P2HP02EN/Jeff+Goldblum+Your+Scientists+Were+so+preoccupied+with+whether+or+not+they+could+they+didn%27t+stop+to+think+if+they+should.jpg){fig-align="center"}


---

## Displaying trend lines: linear regression

```{r}
#| code-line-numbers: "4"
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```


---

## Setup and motivation for linear regression


::: {style="font-size: 85%;"}

Consider an outcome $Y \in \mathbb{R}$ and covariate $X \in \mathbb{R}$

  + We have $n$ observations: $(X_1, Y_1), \dots, (X_n, Y_n)$
  
Purpose of regression is to model $\mathbb{E}[Y | X]$

Consider the case where $X$ takes on discrete values $c_1, \dots, c_k$

Then most straightforward way to estimate $\mathbb{E}[Y | X = c_j]$ is to use the sample mean for subgroup $X_i = c_j$:

$$\hat{\mathbb{E}}[Y|X = c_j] = \frac{1}{N_j} \sum_{i: X_i = c_j} Y_i$$

+ Graphs like side-by-side violin plots, facetted histograms, and overlaid density plots essentially compare $\hat{\mathbb{E}}[Y|X = c_j]$ for different categories

But when $X$ is quantitative, what do we do?

+ Use statistical model to "guess" $\mathbb{E}[Y|X = x]$, even when we don't observe $X = x$

:::

---

## Statistical Model for Linear Regression

Linear regression assumes $Y_i \overset{iid}{\sim} N(\beta_0 + \beta_1 X_i, \sigma^2)$

- $\beta_0$: _intercept_ - population mean outcome when $X = 0$; i.e., $\mathbb{E}[Y | X = 0]$

- $\beta_1$: _slope_ - population mean _change_ in $Y$ when $X$ increases by 1

- $\beta_0$ and $\beta_1$ are parameters that must be estimated

<!--

The assumptions baked into this model are:

1. Normality

2. Equal variance

3. Independent errors

4. Linearity

5. Fixed $X$

-->

---

## Assessing assumptions of linear regression

Linear regression assumes $Y_i \overset{iid}{\sim} N(\beta_0 + \beta_1 X_i, \sigma^2)$

- If this is true, then $Y_i - \hat{Y}_i \overset{iid}{\sim} N(0, \sigma^2)$

. . .

Plot residuals against $\hat{Y}_i$, __residuals vs fit__ plot

- Used to assess linearity, any divergence from mean 0

- Used to assess equal variance, i.e., if $\sigma^2$ is homogenous across predictions/fits $\hat{Y}_i$

. . .

More difficult to assess the independence and fixed $X$ assumptions

- Make these assumptions based on subject-matter knowledge

---

## Residual vs fit plots


```{r}
lin_reg <- lm(body_mass_g ~ flipper_length_mm, data = penguins)

tibble(fits = fitted(lin_reg), residuals = residuals(lin_reg)) |>
  ggplot(aes(x = fits, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")
```


---

## Residual vs fit plots

```{r}
#| code-line-numbers: "5"
tibble(fits = fitted(lin_reg), residuals = residuals(lin_reg)) |>
  ggplot(aes(x = fits, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth()
```

---

## Examples of Residual-vs-Fit Plots




---

## Recap and next steps

::: {style="font-size: 80%;"}

Scatterplots are the most common visual for 2D quantitative variables

+ Many ways to incorporate additional dimensions in scatterplots, e.g., color and shape

Linear regression is by far the most common model for describing the relationship between 2+ quantitative variables

+ Can also: transform the outcome, transform the covariates, do nonparametric "smoothing"

Use graphs to assess linear regression assumptions, i.e., residual diagnostic plots

:::

. . .

::: {style="font-size: 80%;"}

+ **HW4 due Wednesday and you have Lab 6 on Friday**

+ **Graphics critique due Feb 28th!**

+ **Next time**: Inference with Linear Regression

+ Recommended reading: [CW Chapter 12 Visualizing associations among two or more quantitative variables](https://clauswilke.com/dataviz/visualizing-associations.html)

:::

